---
title: "Final Project"
author: "Jordan Kakuyo"
date: "2024-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Analysis for eSC houses of a newer make (>2000s)

## Packages

```{r}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(arrow)
library(caret)
library(party)
library(partykit)
library(Metrics)

houses <-  read_parquet("~/Desktop/Syracuse University /Grad School/Year 1/Intro to Data Science/Final_Project/julyData_middle2 1.parquet")

jmd_1 <- houses

head(houses)
```

```{r}
houses_numeric <- houses
for (col in colnames(houses_numeric)) {
  if (!is.numeric(houses_numeric[[col]])) {
    houses_numeric[[col]] <- as.numeric(as.factor(houses_numeric[[col]]))
  }
}  
  
#data.frame(lapply(houses, function(x) as.numeric(as.character(x))))
matrix <- cor(houses_numeric)
matrix <- as.data.frame(matrix)
matrix

# corrplot(matrix, method="color")
```

## Data Manipulation

```{r}
library(qs)

houses <- houses %>% select(-era) %>% select(-in.hvac_has_shared_system)
houses <- houses[, -1]

houses$hour<-format(houses$date_time, format = "%H:%M:%S")
#houses <- houses %>% select(-date_time) %>% select(-in.county)

for (col in colnames(houses)) {
  if (!is.numeric(houses[[col]])) {
    houses[[col]] <- as.numeric(as.factor(houses[[col]]))
  }
}  


library(qs)
qsave(houses, "houses_data2.qs")



summary(houses)
```
## Training and Testing Sets

```{r}
set.seed(123)

train <- sample(c(TRUE, FALSE), nrow(houses), replace=TRUE, prob=c(0.7,0.3))
train_data <- houses[train, ]  # Training data
test_data <- houses[!train, ]  # Testing data
```

## Decision Tree 2: Chosen model

```{r}
modelTree<- ctree(out.electricity.cooling.energy_consumption ~ ., train_data)


summary(modelTree)

predictions <- predict(modelTree, test_data, type = "response")


mae(test_data$out.electricity.cooling.energy_consumption, predictions)
# 0.1811066
rmse(test_data$out.electricity.cooling.energy_consumption, predictions)
# 0.2857714
R2(predictions, test_data$out.electricity.cooling.energy_consumption)
# 0.5504721
```
```{r}

library(party)
library(rpart)


#str(houses)

#control <- ctree_control(
  #mincriterion = 0.99,  # Require stronger evidence for splits
 # minsplit = 20,        # Require at least 20 observations to attempt a split
 # maxdepth = 5          # Limit the maximum depth of the tree
#)


set.seed(123)

train <- sample(c(TRUE, FALSE), nrow(houses), replace=TRUE, prob=c(0.7,0.3))
train_data <- houses[train, ]  # Training data
test_data <- houses[!train, ]  # Testing data
sampled_data <- train_data[sample(1:nrow(train_data), size = 0.1 * nrow(train_data)), ]


houses <- houses[,1:100000]
qsave(houses, "houses_data3.qs")



modelTree2 <- ctree(out.electricity.cooling.energy_consumption ~ ., data = sampled_data, controls = ctree_control(
    maxdepth = 5,         # Maximum depth of the tree
    mincriterion = 0.99,  # Higher threshold for splits (fewer splits)
    minsplit = 30         # Minimum observations per split
  )
)

predictions <- predict(modelTree2, test_data, type = "response")

mae(test_data$out.electricity.cooling.energy_consumption, predictions)

saveRDS(modelTree2, "modelTree2.rds", compress = TRUE)



object_size(sampled_data) #19.26 

print(modelTree2) #went up to 1211

str(modelTree2, max.level = 1) #Formal class 'BinaryTree' [package "party"] with 10 slots



```






## Model Creation (Logistic Regression)

```{r}
modelLog <- glm(out.electricity.cooling.energy_consumption ~ ., gaussian, train_data)
predictionsLog <- predict(modelLog, test_data, type = "response")
```


```{r}
mae(test_data$out.electricity.cooling.energy_consumption, predictionsLog)
# 0.2300664
rmse(test_data$out.electricity.cooling.energy_consumption, predictionsLog)
# 0.3464799
R2(predictionsLog, test_data$out.electricity.cooling.energy_consumption)
# 0.3379966
```
## Model Creation (Linear)

```{r}
modelLin <- lm(out.electricity.cooling.energy_consumption ~ ., train_data)
predictionsLin <- predict(modelLin, test_data, type = "response")
```

```{r}
mae(test_data$out.electricity.cooling.energy_consumption, predictionsLin)
# 0.23006644
rmse(test_data$out.electricity.cooling.energy_consumption, predictionsLin)
# 0.3464799
R2(predictionsLin, test_data$out.electricity.cooling.energy_consumption)
# 0.33379966
```
## Model Creation (Gradient Boosting)

```{r}
library(gbm)

gbm_model <- gbm(out.electricity.cooling.energy_consumption ~ ., data = train_data, distribution = "gaussian")
predictionsGLM <- predict(gbm_model, test_data, type = "response")
```

```{r}
mae(test_data$out.electricity.cooling.energy_consumption, predictionsGLM)
# 0.2398109
rmse(test_data$out.electricity.cooling.energy_consumption, predictionsGLM)
# 0.3646004
R2(predictionsGLM, test_data$out.electricity.cooling.energy_consumption)
# 0.3159185
```


## Prepping DF for classification

```{r}
houses_class <- houses
houses_class$out.electricity.cooling.energy_consumption <- cut(
  houses_class$out.electricity.cooling.energy_consumption,
  breaks = c(0, 2.6, 5.2, Inf),  
  labels = c("Low", "Medium", "High")
)

View(houses_class)
```


## Test and Training Data

```{r}
set.seed(123)

train_class <- sample(c(TRUE, FALSE), nrow(houses_class), replace=TRUE, prob=c(0.7,0.3))
train_data_class <- houses_class[train_class, ]  # Training data
test_data_class <- houses_class[!train_class, ]  # Testing data
```


## Decison Tree

```{r}
#modelTreeClass <- ctree(out.electricity.cooling.energy_consumption ~ ., data = train_data_class)

#nrow(train_data_class)
#length(train_data_class$out.electricity.cooling.energy_consumption)

# Step 4: Make Predictions
#predictions <- predict(modelTreeClass, test_data_class, type = "response")

# Step 5: Evaluate Classification Metrics
# Confusion Matrix
#conf_matrix <- table(Predicted = predictions, Actual = test_data_class$out.electricity.cooling.energy_consumption)
#print(conf_matrix)

# Accuracy
#accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
#cat("Accuracy:", accuracy, "\n")
```
## Visualizations 

## Square Feet
```{r}
ggplot(houses, aes(x = in.sqft, y = out.electricity.cooling.energy_consumption)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Stories
```{r}
ggplot(houses, aes(x = in.geometry_stories, y = out.electricity.cooling.energy_consumption)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Bedrooms
```{r}
ggplot(houses, aes(x = in.bedrooms, y = out.electricity.cooling.energy_consumption)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Relative Humidity
```{r}
ggplot(houses, aes(x = `Relative Humidity [%]`, y = out.electricity.cooling.energy_consumption)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Temperature
```{r}
ggplot(houses, aes(x = `Dry Bulb Temperature [°C]`, y = out.electricity.cooling.energy_consumption)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Income
```{r}
ggplot(houses, aes(x = in.income, y = out.electricity.cooling.energy_consumption)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Income + Infiltration
```{r}
ggplot(housesV, aes(x = in.income, fill = in.infiltration)) +
  geom_bar(position = "stack") +
  theme_minimal() +
  labs(x = "Income", y = "Count", fill = "Infiltration Type",
       title = "Stacked Bar Plot")
```

##
```{r}
data_percent2 <- housesV %>%
  count(in.infiltration ,in.geometry_wall_type) %>%
  group_by(in.infiltration) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(data_percent2, aes(x = in.infiltration, y = n, fill = in.geometry_wall_type)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),  
            color = "white", size = 3) +  
  theme_minimal() +
  labs(x = "Infiltration Type", y = "Count", fill = "Wall Type", 
       title = "Stacked Bar Plot with Percentages")
```
## Usage + Wall Type
```{r}
ggplot(housesV, aes(x = in.geometry_wall_type, fill = in.usage_level)) +
  geom_bar(position = "stack") +
  theme_minimal() +
  labs(x = "Wall Type", y = "Count", fill = "Infiltration Type",
       title = "Stacked Bar Plot")
```
```{r}
data_percent <- housesV %>%
  count(in.geometry_wall_type, in.usage_level) %>%
  group_by(in.geometry_wall_type) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(data_percent, aes(x = in.geometry_wall_type, y = n, fill = in.usage_level)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),  
            color = "white", size = 4) +  
  theme_minimal() +
  labs(x = "Wall Type", y = "Relative Usage Level", fill = "Category 2", 
       title = "Stacked Bar Plot with Percentages")
```
```{r}
data_percent <- housesV %>%
  count(in.geometry_wall_type, in.usage_level) %>%
  group_by(in.geometry_wall_type) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(data_percent, aes(x = in.geometry_wall_type, y = n, fill = in.usage_level)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),  
            color = "white", size = 4) +  
  scale_y_continuous(limits = c(0, 10000), expand = c(0, 0)) +
  theme_minimal() +
  labs(x = "Wall Type", y = "Relative Usage Level", fill = "Infiltration Type", 
       title = "Stacked Bar Plot with Percentages")
```

```{r}
data_percent <- housesV %>%
  count(in.geometry_wall_type, in.usage_level) %>%
  group_by(in.geometry_wall_type) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(data_percent, aes(x = in.geometry_wall_type, y = n, fill = in.usage_level)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),  
            color = "white", size = 4) +  
  scale_y_continuous(limits = c(0, 50000), expand = c(0, 0)) +
  theme_minimal() +
  labs(x = "Wall Type", y = "Relative Usage Level", fill = "Infiltration Type", 
       title = "Stacked Bar Plot with Percentages")






## Making New Data Frame to Test
```

```{r}
housesFive <- houses
housesFive$`Dry Bulb Temperature [°C]` <- housesFive$`Dry Bulb Temperature [°C]` + 5
```

## Using Model to Make Predictions
```{r}
predictionsFive <- predict(modelTree, housesFive, type = "response") 

head(predictionsFive)
```


## Add Predictions to the Table
```{r}
housesFive$Energy_consumption_prediction <- predictionsFive
housesVFive <- housesV
housesVFive$Energy_consumption_prediction <- predictionsFive

average_consumption <- housesVFive %>%
  group_by(in.county) %>%
  summarise(avg_energy_consumption = mean(Energy_consumption_prediction, na.rm = TRUE))



```
```
